---
title: "Transforms between yakuba and MANC"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This article documents the registrations used to map the *D. yakuba* VNC
onto the *D. melanogaster* MANC template, and onto a symmetrised version of
itself for left-right mirroring. It also explains a coordinate offset that was
found and corrected in September 2026.

## Summary

- `xform_brain(x, sample = "yakuba", reference = "MANC")` and
  `xform_dyak2manc(x)` use the default `"tps1000"` registration.
- `mirror_dyak()`, `symmetric_dyak()` and `dyak_lr_position()` use the
  `yakubasym` registration.
- Both registrations were computed in a coordinate frame offset from current
  yakuba data by about (6.5, 7.1, 8.0) µm. Points are now shifted by
  (-6.5, -7.1, -8.0) µm before either registration is applied.
- Two alternative yakuba to MANC registrations are available with
  `xform_dyak2manc(x, method = "manual")` and `method = "ngscene"`. The old
  registration without the shift is `method = "tps1000_unshifted"`.

```{r}
library(yakuba)
mdn <- read_dyak_neurons("MDN")
mdn.manc <- xform_dyak2manc(mdn)                     # default, = xform_brain
mdn.manual <- xform_dyak2manc(mdn, method = "manual")
mdn.ng <- xform_dyak2manc(mdn, method = "ngscene")   # needs malecns
mdn.m <- mirror_dyak(mdn)
```

## Registrations

| Method | Landmarks | Source | Space |
|---|---|---|---|
| `tps1000` (default) | 1000 | Sebastian Cachero | yakuba → MANC, via yakSurf frame |
| `manual` | 100 | Hiroshi Shiozaki (placed by hand) | yakuba → MANC |
| `ngscene` | 21 | clio neuroglancer scene (pointA/pointB) | yakuba → male CNS, then `malecnsum` → MANC |
| `yakubasym` | 1000 | Sebastian Cachero | yakuba → symmetrised yakuba, via yakSurf frame |

All registrations are thin-plate splines in microns (`yakubaum`). Inverses
are computed by swapping the landmark sets, so round trips are approximate
(~0.3 µm for `tps1000`, a few µm for `manual`).

The two 1000-point registrations were computed against *yakSurf*, a
neuropil surface that Sebastian built from the yakuba synapse cloud available
at the time (`malevnc/data-raw/yakSurf_lowRes.rds`).

## The offset

### Symptoms

Before the correction:

- `mirror_dyak()` did not map left neuropils onto their right partners. After
  mirroring each (L) neuprint ROI, a median 10.5 µm shift in X was still
  needed to match the corresponding (R) ROI. The same check on MANC with
  `malevnc::mirror_manc()` gives ~0.
- Yakuba neurons mapped to MANC were displaced from their matched MANC
  partners, by the same amount in the same direction on both sides of the
  VNC. For MDN, DNg13 and DNa02 the median offset was (4.7, -1.8, 1.5) µm.
  Shared neuropil ROIs were displaced by a median (7.4, -2.1, 10.1) µm.

A bias shared by both registrations, and constant across the VNC, pointed to
a problem with the input frame rather than the registrations themselves.

### Measurement

The clio neuroglancer scene for yakuba includes a `neuropil-shell` layer. This
is shipped with the package as `yakuba_neuropil_shell`, along with the
`vnc-shell` layer as `yakuba_vnc_shell`. yakSurf matches the neuropil shell
closely after a pure translation, found with translation-only ICP (see
`data-raw/yakuba_surf_offset.R`):

- **Translation:** (6.45, 7.09, 7.92) µm, from yakSurf to the full-resolution
  shell.
- **Fit:** the median surface distance falls from 5.3 to 1.1 µm, which is
  about the precision of meshing at 0.5-2 µm voxels.
- **Not a flip:** all eight combinations of axis flips were tried, and only
  the unflipped surface fits. The best flip, X, leaves 5.7 µm, which is about
  what you get mirroring the shell onto itself.
- **Not scale or rotation:** a similarity fit gives scale 1.000 and 0°
  rotation. Fitting each quarter of the VNC along the AP axis separately gives
  the same translation within ~1 µm.
- **Not the vnc-shell:** the same fit against the outer `vnc-shell` is poor
  (8.3 µm), so yakSurf is a neuropil surface.

The offset is not a whole number of voxels at any plausible grid size, and it
does not match the [0, 122, 896] voxel shift Stuart Berg applied to the
yakuba volume. It most likely comes from padding or cropping of the volume
the original synapse cloud was taken from. The package uses the value
re-derived against the decimated shell, rounded to 0.1 µm:
`yakuba_surf_offset = c(-6.5, -7.1, -8.0)`. The precision is about ±1 µm per
axis.

### Validation

The offset was measured from the surfaces alone, with no neuron data. It was
then checked against neuprint ROI meshes and three matched descending neuron
types, using `data-raw/yakuba_xform_validation.R`:

| Registration | DN mean distance (µm) | Median ROI shift vs MANC (µm) |
|---|---|---|
| tps1000_unshifted (old) | 8.9 | (7.4, -2.1, 10.1) |
| **tps1000 (default)** | **5.6** | **(0.6, 0.2, 0.2)** |
| manual | 6.4 | (0.9, -0.3, -1.2) |
| ngscene | 7.7 | (-2.0, 2.2, 1.6) |

- **DN mean distance** is the mean distance from each transformed yakuba MDN,
  DNg13 and DNa02 to the closest MANC neuron of the same type.
- **Median ROI shift** is the translation still needed to line up each shared
  ROI after transformation. For a well-registered pair it should be about 0.
- **Mirroring:** the remaining X mismatch after `mirror_dyak()` of (L) onto
  (R) ROIs fell from 10.5 to 0.2 µm.
- **Independent fits:** fitting the best pre-shift directly to the DNs gave
  about (-6, -5.5, -4.8) µm, and to the ROIs about (-7, -7.8, -7.8) µm. Both
  are consistent with the surface-derived value.
- **malecns:** results for the route to the male CNS improve by the same
  amount, since it passes through MANC.

### Left-right assignment of somata

`dyak_lr_position()` was also checked against the neuprint `somaSide`
annotation for about 12,000 intrinsic neurons. A positive value means the fly's
right. `malevnc::manc_lr_position()` on MANC intrinsic neurons is the
benchmark.

| | n (L/R) | Sign matches `somaSide` | Median L / R (µm) | Median of M somata (µm) |
|---|---|---|---|---|
| yakuba, no offset (old) | 12145 | 96.6% | -97.2 / 75.0 | -4.3 |
| **yakuba (current)** | 12145 | **97.9%** | **-86.5 / 85.9** | 5.7 |
| MANC | 12706 | 97.2% | -88.8 / 94.6 | -0.9 |

- **Symmetry:** without the offset, the L and R distributions were off-centre by
  about 11 µm. With it they are symmetric.
- **Errors:** most wrongly assigned somata are near the midline, with a median
  |value| of 5 µm. The error rate is similar to MANC, so it is about the limit
  of the annotation and of soma positions near the midline.
- **Midline somata:** somata annotated `M` scatter around 0, with a MAD of about
  17 µm.

What remains after the correction (for example ~5.6 µm for the DNs) reflects
the accuracy of the registrations and real anatomical differences between the
species. It is scatter, not a consistent offset.

## Reproducing

- `data-raw/yakuba_shells.R`: downloads and decimates the clio shells.
- `data-raw/yakuba_surf_offset.R`: measures the offset and runs the checks
  above. It needs `yakSurf_lowRes.rds` from the malevnc repository.
- `data-raw/yakuba_malecns_ngscene_tps.R`: builds the `ngscene` registration.
- `data-raw/yakuba_xform_validation.R`: produces the validation tables above,
  including the soma check.
  It needs neuprint access to yakuba and MANC, and takes about a minute.
